---
title: "Data Visualization"
author: "Adejumo Ridwan Suleiman"
date: "`r Sys.Date()`"
output: html_document
---

# Pie Chart
pie(X, Labels, Radius, Main, Col, Clockwise)  

- X is a vector that contains the numeric values used in the pie chart.
- Labels are used to give the description to the slices.
- Radius describes the radius of the pie chart.
- Main describes the title of the chart.
- Col defines the color palette.
- Clockwise is a logical value that indicates the clockwise or anti-clockwise direction in which slices are drawn.

```{r}
# Creating data for the graph.  
x <- c(20, 65, 15, 50)  
labels <- c("India", "America", "Shri Lanka", "Nepal")  

# Plotting the chart.  
pie(x,labels)  
```

```{r}
# Creating data for the graph.  
x <- c(20, 65, 15, 50)  
labels <- c("India", "America", "Shri Lanka", "Nepal")  
  
# Plotting the chart.  
pie(x,labels,main="Country Pie chart",col=rainbow(length(x)))  
```
```{r}
# Creating data for the graph.  
x <- c(20, 65, 15, 50)  
labels <- c("India", "America", "Shri Lanka", "Nepal")  
pie_percent<- round(100*x/sum(x), 1)  

# Plotting the chart.  
pie(x, 
    labels = pie_percent, 
    main = "Country Pie Chart",
    col = rainbow(length(x))) 

legend("topright", 
       c("India", "America", "Shri Lanka", "Nepal"), 
       cex = 0.8,
       fill = rainbow(length(x)))  
```

# Bar Plot
barplot(h,x,y,main, names.arg,col)  

```{r}
# Creating the data for Bar chart  
H<- c(12,35,54,3,41)  

# Plotting the bar chart   
barplot(H)  

head(mtcars)

```

```{r}
# Creating the data for Bar chart  
H <- c(12,35,54,3,41)  
M<- c("Feb","Mar","Apr","May","Jun")  
  

# Plotting the bar chart   
barplot(H,
        names.arg=M,
        xlab="Month",
        ylab="Revenue",
        col="Green",  
        main="Revenue Bar chart",
        border="red") 


```

# Scatterplot

```{r}

```

