---
title: "R Data Structures"
author: "Adejumo Ridwan Suleiman"
date: '`r Sys.Date()`'
output: html_document
---

## Vectors
- A vector is a type of data type in R that combines elements of the same data types
- The c() function is use to create a vector
- Let's say we have a basket containing fruits, we can use a vector to represent it below
```{r}
basket <- c("mango", "orange", "banana", "pineapple")
basket
```
- When we put a numerical value in the above vector, the vector converts it into a string or charactert element. It is no more a numerical value and this is due to the fact that all elements of a vector must be of the same data type.
```{r}
basket <- c(4, "orange", "banana", "pineapple")
basket
```
- We can also combine numerical elements using the c() function or we can use a colon (:) if the element are in a sequence.
```{r}
die = c(1,2,3,4,5,6)
#or
die = 1:6

```

- To know the lenght of our vector we use the length() function
```{r}
length(basket)
length(die)
```
- We can access the elements in a vector using indexing, each element in a vector has a unique position called index. To access this element we call it's position. The first item has an index 1, the next 2 and so on.
- To call more than one elements, we use the c() also.
```{r}
#call a single element
basket[2]
die[3]

# call multiple elements
basket[c(3,4)]
```
- You can also change an item in a position by calling that element in that position using it's index and assigning it to a new value.
```{r}
basket[1] <- "apple"
basket
```
## Lists
- Unlike the vector, the list is used to combine elements of various data types.
- A list is created using the list function
```{r}
die_and_basket <- list("orange",1,2,"apple","lemon",3,4,5,"pear",6)
die_and_basket
```
- A list can be accessed also with indexing using the square brackets. The first item has an index 1, the next 2 and so on.
```{r}
die_and_basket[2]
```
- To change an item value in a list, you refer to the index number and assign it a new value.
```{r}
die_and_basket[3] <- "pawpaw"
die_and_basket
```
- You can get the length of a list using the length() function.
```{r}
length(die_and_basket)
```
- To add an item to a list, you use the append function
```{r}
append(die_and_basket, "guava")
```
- To remove an item in a list, you use a negative indexing and ovewrite the previous list with the new list
```{r}
die_and_basket
new_die_and_basket <- die_and_basket[-3]
new_die_and_basket
```
- You can also loop through a list using a for loop
```{r}
for (i in die_and_basket){
  print(i)
}
```
- Two lists can be joined using the c() function.
```{r}
list_1 <- list(1,"boy")
list_2 <- list(4,"girl")
list_3 <- c(list_1,list_2)
list_3
```

## Matrices








