---
title: "R Data Structures"
author: "Adejumo Ridwan Suleiman"
date: '`r Sys.Date()`'
output: html_document
---

## Vectors
- A vector is a type of data type in R that combines elements of the same data types
- The c() function is use to create a vector
- Let's say we have a basket containing fruits, we can use a vector to represent it below
```{r}
basket <- c("mango", "orange", "banana", "pineapple")
basket
```
- When we put a numerical value in the above vector, the vector converts it into a string or charactert element. It is no more a numerical value and this is due to the fact that all elements of a vector must be of the same data type.
```{r}
basket <- c(4, "orange", "banana", "pineapple")
basket
```
- We can also combine numerical elements using the c() function or we can use a colon (:) if the element are in a sequence.
```{r}
die = c(1,2,3,4,5,6)
#or
die = 1:6

```

- To know the lenght of our vector we use the length() function
```{r}
length(basket)
length(die)
```
- We can access the elements in a vector using indexing, each element in a vector has a unique position called index. To access this element we call it's position. The first item has an index 1, the next 2 and so on.
- To call more than one elements, we use the c() also.
```{r}
#call a single element
basket[2]
die[3]

# call multiple elements
basket[c(3,4)]
```
- You can also change an item in a position by calling that element in that position using it's index and assigning it to a new value.
```{r}
basket[1] <- "apple"
basket
```
## Lists
- Unlike the vector, the list is used to combine elements of various data types.
- A list is created using the list function
```{r}
die_and_basket <- list("orange",1,2,"apple","lemon",3,4,5,"pear",6)
die_and_basket
```
- A list can be accessed also with indexing using the square brackets. The first item has an index 1, the next 2 and so on.
```{r}
die_and_basket[2]
```
- To change an item value in a list, you refer to the index number and assign it a new value.
```{r}
die_and_basket[3] <- "pawpaw"
die_and_basket
```
- You can get the length of a list using the length() function.
```{r}
length(die_and_basket)
```
- To add an item to a list, you use the append function
```{r}
append(die_and_basket, "guava")
```
- To remove an item in a list, you use a negative indexing and ovewrite the previous list with the new list
```{r}
die_and_basket
new_die_and_basket <- die_and_basket[-3]
new_die_and_basket
```
- You can also loop through a list using a for loop
```{r}
for (i in die_and_basket){
  print(i)
}
```
- Two lists can be joined using the c() function.
```{r}
list_1 <- list(1,"boy")
list_2 <- list(4,"girl")
list_3 <- c(list_1,list_2)
list_3
```

## Matrices
- A matrix is a two dimensional data struture with rows and columns.
- A matrix can be generated using the matrix() function
```{r}
my_matrix <- matrix(c(1,2,3,4,5,6), nrow = 3, ncol = 2)
my_matrix
```
- Matrix can also be created with strings or characters.
```{r}
my_second_matrix <- matrix(c("boy","man","girl","woman"), nrow = 2, ncol = 2)
my_second_matrix
```
- To access the items in a matrix, you use the square brackets.
- The first number in the bracket specifies the row position while the second number in the bracket specifies the column position.
```{r}
#to access the element in row 1 and column 2 of my_matrix
my_matrix[1,2]

# to access the element in row 2 and column of 1 of my_second_matrix
my_second_matrix[2,1]

# to access the whole row in column 1 of my_matrix
my_matrix[,1]

# to access the whole column in row 1 of my_matrix
my_matrix[1,]

# to access more than one row
my_matrix[c(1,2),]

# to access more than one column and row
my_matrix[c(1,2),c(1,2)]
```

- You can add more rows and columns to a matrix using the cbind() for column addition and rbind() for row addition column.
```{r}
#add a new column
my_matrix <- cbind(my_matrix, c("animal","men","fairies"))
my_matrix

#add a new row
my_matrix <- rbind(my_matrix, c("car", "bike", "van"))
my_matrix
```

- 
```{r}
# remove rows and columns
my_matrix <- my_matrix[-c(1),-c(3)]
my_matrix

#Check if an item is in the matrix using %in%
"car" %in% my_matrix

#Check for the dimension of the rows and columns
dim(my_matrix)

#check for the lenght of a matrix
length(my_matrix)

```
## Data Frames
- Data frame are data structures used to display data in a tabular format.
- In a data frame contains columns with different data types but each column must carry the same type of data.
- A data frame is created using the data.frame() function.
```{r}
my_dataFrame <- data.frame(
  Names = c("Jacob", "Mary","Seth"),
  Age = c(20,14,23),
  Sex = c("Male","Female","Male"),
  Married = c(TRUE, FALSE, TRUE)
)
my_dataFrame

```
- To summarize the data, we use the summary() function
```{r}
summary(my_dataFrame)
my_dataFrame[[2]]
```
- To Access columns in the dataframe we use the [] brackets, [[]] brackets or $ sign
```{r}
my_dataFrame["Age"]
my_dataFrame[["Age"]] #returns the column as a vector
my_dataFrame$Age

```
- To add rows and columns we can use the rbind() and cbind() function respectively just like in the matrix.
```{r}
# add rows
new_dataframe <- rbind(my_dataFrame, c("Alice", 18, "Female", TRUE))
new_dataframe

another_dataframe <- cbind(my_dataFrame, Education = c("Bachelors","HighSchool","PhD"))
another_dataframe
```
- To remove rows and columns we using negative indexing
```{r}
#remove first row and column
remove <- my_dataFrame[-1,-1]
remove

```
- Amount of rows and columns can be checked using the dim() function or nrow() and ncol() for number of rows and columns respectively
```{r}
dim(my_dataFrame)
nrow(my_dataFrame)
ncol(my_dataFrame)
```
- you can combine two data frames together either by rows using the rbind() function or by columns using the cbind() function.
```{r}
#combine by column
my_dataFrame <- data.frame(
  Names = c("Jacob", "Mary","Seth"),
  Age = c(20,14,23)
)
second_dataframe <- data.frame(
  Sex = c("Male","Female","Male"),
  Married = c(TRUE, FALSE, TRUE)
)
third_dataframe <- cbind(my_dataFrame, second_dataframe)
third_dataframe
```
```{r}
#combine by row
myrow_dataFrame <- data.frame(
  Names = c("Jacob", "Mary","Seth"),
  Age = c(20,14,23),
  Sex = c("Male","Female","Male"),
  Married = c(TRUE, FALSE, TRUE)
)
second_row_dataframe <- data.frame(
  Names = c("John", "Chris","David"),
  Age = c(13,23,22),
  Sex = c("Male","Female","Male"),
  Married = c(TRUE, FALSE, TRUE)
)

third_row_dataframe <- rbind(myrow_dataFrame, second_row_dataframe)
third_row_dataframe
```

## Factors
- Factors are data type expressed categorically e.g
* Sex: Male or Female
* Temperature Level: High, Medium and Low
* Size: Big or Small
- A factor is created using the factor() function
```{r}
Temperature <- factor(c("High","High","Low","Medium","Low","Medium","High"))
Temperature
```
- Above we have 7 elements but 3 levels.
- To access the levels of a factor we use the levels function
```{r}
levels(Temperature)
```
- To assign levels we can use the levels arguments in the function.
```{r}
Temperature <- factor(c("High","High","Low","Medium","Low","Medium","High"), levels = c("High","Medium","Low"))
Temperature
```









 